Management induced changes of soil organic carbon on global croplands

### 2021-08-09 SOCbudgetPaper/SOCbudget.Rmd
remove copyright statement fix typo in reference
udate paper draft - intro, abstract, small changes in methods

---
title: Management induced changes of soil organic carbon on global croplands 
journal: "`r rticles::copernicus_journal_abbreviations(journal_name = 'communication')`"
author:
  - given_name: Kristine
    surname: Karstens
    affiliation: "1, 3"
    email: kristine.karstens@pik-potsdam.de
    orcid: https://orcid.org/0000-0002-5860-9789
    corresponding: true
  - given_name: Benjamin Leon
    surname: Bodirsky
    affiliation: 1
    email: bodirsky@pik-potsdam.de
    orcid: https://orcid.org/0000-0002-8242-6712
  - given_name: Jan Philipp
    surname: Dietrich
    affiliation: 1 
    orcid: https://orcid.org/0000-0002-4309-6431
  - given_name: Marta
    surname: Dondini
    affiliation: 2
  - given_name: Jens
    surname: Heinke
    affiliation: 1
  - given_name: Matthias
    surname: Kuhnert
    affiliation: 2
  - given_name: Christoph
    surname: Müller
    affiliation: 1 
    orcid: https://orcid.org/0000-0002-9491-3550
  - given_name: Susanne
    surname: Rolinski
    affiliation: 1
  - given_name: Pete
    surname: Smith
    affiliation: 2
    orcid: https://orcid.org/0000-0002-3784-1124
  - given_name: Isabelle
    surname: Weindl
    affiliation: 1
    orcid: https://orcid.org/0000-0002-7651-6930
  - given_name: Hermann
    surname: Lotze-Campen
    affiliation: "1, 3"
  - given_name: Alexander
    surname: Popp
    affiliation: 1
affiliation:
  - code: 1
    address: Potsdam Institute for Climate Impact Research (PIK), Member of the Leibniz Association, P.O. Box 60 12 03, 14412 Potsdam, Germany
  - code: 2
    address: Institute of Biological & Environmental Sciences, University of Aberdeen, Aberdeen, UK
  - code: 3
    address: Humboldt-Universität zu Berlin, Department of Agricultural Economics, Unter den Linden 6, 10099 Berlin, Germany
abstract: |
  Soil organic carbon (SOC) is one of the largest terrestrial carbon stocks on Earth. The first meter of the Earths soils profile stores three times as much carbon as the vegetation and twice the amount of C in the atmosphere. SOC has been depleted by anthropogenic land cover change and agricultural management. However, the latter has so far not been well represented in global carbon stock assessments. While SOC models often simulate detailed biochemical processes that lead to the accumulation and decay of SOC, the management decisions driving these biophysical processes are still little investigated at the global scale. Here we develop a spatial explicit data set for agricultural management on cropland, considering crop production levels, residue returning rates, manure application, and the adoption of irrigation and tillage practices. We combine it with the IPCC Tier 2 steady-state soil model to create a half-degree resolution data set of SOC stocks and SOC stock changes for the first 30 cm of mineral soils. We estimate that due to arable farming, soils have lost around 26 GtC relative to a counterfactual natural state in 1975. Yet, within the period 1975--2010 this SOC debt has been decreasing again by a net quantity of 4 Gt SOC, which can be mainly traced back to an increased input of C in crop residues due to higher crop productivity. We also find that SOC is very sensitive to management decisions such as residue returning indicating the necessity to incorporate better management data in soil model simulations.

bibliography: SOCbudget.bib
running:
  title: R Markdown Template for Copernicus
  author: Nüst et al.
competinginterests: |
  The authors declare no competing interests.
# OPTIONAL:
algorithms: true
# See https://publications.copernicus.org/for_authors/licence_and_copyright.html, normally used for transferring the copyright, if needed.
copyrightstatement: |
  The author's copyright for this publication is transferred to institution/company.
availability:
  codedata: |
   We compile calculations as open-source R packages available at github.com/pik-piam/mrcommons \citep{bodirsky_mrcommons_2020} for the management related functions, github.com/pik-piam/mrsoil \citep{karstens_mrsoil_2020} for soil dynamic related functions and  github.com/pik-piam/mrvalidation \citep{bodirsky_mrvalidation_2020} for validation data. All libraries are based on the MADRaT package at github.com/pik-piam/madrat \citep{dietrich_madrat_2020}, a framework which aims to improve reproducibility and transparency in data processing. Model results including C input data are accessable under https://doi.org/10.5281/zenodo.4320663 \citep{karstens_model_2020}. Software code for paper and result prepartion can be found under www.github.com/k4rst3ns/historicalsocmanegement.
authorcontribution: |
  KK, BLB and AP designed the study and the model idea. KK wrote the code build on work of BLB, IW. JPD revised and improved the model code. CM, JH and SR provided the LPJmL simulation data. KK wrote the paper with important contributions of BLB and CM. MK, JS, SR and IW provided extensive feedback to outline of the study. All authors discussed the results and commented on the manuscript.

>>>>

---
title: Management induced changes of soil organic carbon on global croplands 
journal: "`r rticles::copernicus_journal_abbreviations(journal_name = 'communication')`"
author:
  - given_name: Kristine
    surname: Karstens
    affiliation: "1, 3"
    email: kristine.karstens@pik-potsdam.de
    orcid: https://orcid.org/0000-0002-5860-9789
    corresponding: true
  - given_name: Benjamin Leon
    surname: Bodirsky
    affiliation: 1
    email: bodirsky@pik-potsdam.de
    orcid: https://orcid.org/0000-0002-8242-6712
  - given_name: Jan Philipp
    surname: Dietrich
    affiliation: 1 
    orcid: https://orcid.org/0000-0002-4309-6431
  - given_name: Marta
    surname: Dondini
    affiliation: 2
  - given_name: Jens
    surname: Heinke
    affiliation: 1
  - given_name: Matthias
    surname: Kuhnert
    affiliation: 2
  - given_name: Christoph
    surname: Müller
    affiliation: 1 
    orcid: https://orcid.org/0000-0002-9491-3550
  - given_name: Susanne
    surname: Rolinski
    affiliation: 1
  - given_name: Pete
    surname: Smith
    affiliation: 2
    orcid: https://orcid.org/0000-0002-3784-1124
  - given_name: Isabelle
    surname: Weindl
    affiliation: 1
    orcid: https://orcid.org/0000-0002-7651-6930
  - given_name: Hermann
    surname: Lotze-Campen
    affiliation: "1, 3"
  - given_name: Alexander
    surname: Popp
    affiliation: 1
affiliation:
  - code: 1
    address: Potsdam Institute for Climate Impact Research (PIK), Member of the Leibniz Association, P.O. Box 60 12 03, 14412 Potsdam, Germany
  - code: 2
    address: Institute of Biological & Environmental Sciences, University of Aberdeen, Aberdeen, UK
  - code: 3
    address: Humboldt-Universität zu Berlin, Department of Agricultural Economics, Unter den Linden 6, 10099 Berlin, Germany
abstract: |
  Soil organic carbon (SOC) is one of the largest terrestrial carbon stocks on Earth. The first meter of the Earths soils profile stores three times as much carbon as the vegetation and twice the amount of C in the atmosphere. SOC has been depleted by anthropogenic land cover change and agricultural management. However, the latter has so far not been well represented in global carbon stock assessments. While SOC models often simulate detailed biochemical processes that lead to the accumulation and decay of SOC, the management decisions driving these biophysical processes are still little investigated at the global scale. Here we develop a spatial explicit data set for agricultural management on cropland, considering crop production levels, residue returning rates, manure application, and the adoption of irrigation and tillage practices. We combine it with as reduced-complexity model based on the IPCC Tier 2 method to create a half-degree resolution data set of SOC stocks and SOC stock changes for the first 30 cm of mineral soils. We estimate that due to arable farming, soils have lost around 32 GtC relative to a counterfactual natural state in 1975. Within the period 1975--2010 this SOC debt continued to increasing by 5 GtC ($0.14\unit{GtC yr^{-1}}$) to around 37 GtC. However, accounting for historical management rather than assuming constant management conditions led to 2 GtC less $0.06\unit{GtC yr^{-1}}$  emissions. We also find that SOC is most sensitive to management decisions concern soil inputs such as residue returning, indicating that biomass increasing SOC sequestration techniques might be (the more) promising negative emission techniques (NET). Using the here proposed reduced-complexity SOC model it might be possible to track these management-driven SOC gains within integrated (land-use) assessment modeling accounting dynamically for their socio-economic NET potential. 
bibliography: SOCbudget.bib
running:
  title: R Markdown Template for Copernicus
  author: Nüst et al.
competinginterests: |
  The authors declare no competing interests.
# OPTIONAL:
algorithms: true
# See https://publications.copernicus.org/for_authors/licence_and_copyright.html, normally used for transferring the copyright, if needed.
# copyrightstatement: |
#   The author's copyright for this publication is transferred to Potsdam Institute for Climate Impact Research (PIK).
availability:
  codedata: |
   We compile calculations as open-source R packages available at github.com/pik-piam/mrcommons \citep{bodirsky_mrcommons_2020} for the management related functions, github.com/pik-piam/mrsoil \citep{karstens_mrsoil_2020} for soil dynamic related functions and  github.com/pik-piam/mrvalidation \citep{bodirsky_mrvalidation_2020} for validation data. All libraries are based on the MADRaT package at github.com/pik-piam/madrat \citep{dietrich_madrat_2020}, a framework which aims to improve reproducibility and transparency in data processing. Model results including C input data are accessable under https://doi.org/10.5281/zenodo.4320663 \citep{karstens_model_2020}. Software code for paper and result prepartion can be found under www.github.com/k4rst3ns/historicalsocmanegement.
authorcontribution: |
  KK, BLB and AP designed the study and the model idea. KK wrote the code build on work of BLB, IW. JPD revised and improved the model code. CM, JH and SR provided the LPJmL simulation data. KK wrote the paper with important contributions of BLB and CM. MK, JS, SR and IW provided extensive feedback to outline of the study. All authors discussed the results and commented on the manuscript.

### 2021-08-09 SOCbudgetPaper/introduction.Rmd
udate paper draft - intro, abstract, small changes in methods

Natural properties like climatic, biophysical, and landscape characteristics clearly play the most important roles to determine SOC variations over space and time. Recent studies have focused on the evaluation of total SOC stocks of the world as well as on the spatial disaggregation of soil properties such as SOC content [@batjes_harmonized_2016; @hengl_soilgrids250m_2017; @fao_global_2018]. However, these studies often do not include human interventions, like land cover change and agricultural management, in their analysis. Compared to climatic and geological drivers, they alter terrestrial carbon pools over much shorter time scales and are currently one of most dominant drivers of SOC changes on managed land.

The anthropogenic impact can be measured by the SOC debt (also referred to as SOC component of land-use change emissions), which is the amount of organic carbon soils have lost under cultivation compared to a potential natural vegetated state. Sanderman et al. [-@sanderman_soil_2017] identified the anthropogenic SOC debt for the first meter of the soil profile due to land cover change at around 116 GtC (37 GtC for the first 30 cm), compared to previous estimates of 60--130 GtC for the first meter [@lal_world_2001].

Global assessments of the carbon cycle via dynamic global vegetation models (DGVMs) and Earth System Models (ESMs) or bookkeeping models (BKMs) have analyzed SOC losses as part of a comprehensive evaluation of the global carbon budget and land-use change (LUC) emissions [@friedlingstein_global_2019]. While providing estimates of the magnitude of SOC losses due to land cover change, most models lack a detailed consideration of agricultural management. Earlier DGVM and ESM based assessments only considered changes in land cover, but ignored the removal of biomass at harvest [@strassmann_simulating_2008; @betts_climate_2015]. BKMs are designed to estimate LUC related emissons but ignore changes in SOC due to climate change, CO~2~ fertilization, and agricultural management [@friedlingstein_global_2019; @houghton_carbon_2012; @hansis_relevance_2015].

@@

Field-scale models [@del_grosso_simulated_2001, @coleman_simulating_1997, @smith_estimating_2010, @taghizadeh-toosi_c-tool_2014] are able to better account for historic agricultural management if detailed information on crop yield levels, fertilizer inputs and various other on-farm measures is available for the studied sites. However, due to the lack of comprehensive global management data as input to these models, scaling up to the global domain remains a complex challenge.

The objective of our study is to provide the first global, spatially explicit SOC and SOC debt map that considers spatially explicit and time-variant historic agricultural management. To achieve this objective we create and provide a comprehensive data set of the global gridded management time series data, including crop production levels, residue returning rates, manure application, and the adoption of irrigation and tillage practices. We simulate SOC stocks, dynamics and the SOC debt for 1975--2010. Using a scenario analysis, we decompose the contribution of different management activities.

>>>>

Natural properties like climatic, biophysical, and landscape characteristics clearly play the most important roles to determine SOC variations over space and time. Recent studies have focused on the evaluation of total SOC stocks of the world as well as on the spatial disaggregation of soil properties such as SOC content [@batjes_harmonized_2016; @hengl_soilgrids250m_2017; @fao_global_2018]. However, these studies often do not include human interventions, like land cover change and agricultural management, in their analysis. Compared to climatic and geological drivers, they alter terrestrial carbon pools over much shorter time scales and are currently one of most dominant drivers of SOC changes on managed land (cite more book keeping models??).

The anthropogenic impact can be measured by the SOC debt (also referred to as SOC component of land-use change emissions (find corresponding definition of LULUC emission in Pongratz, 2014), which is the amount of organic carbon soils have lost under cultivation compared to a potential natural vegetated state. Sanderman et al. [-@sanderman_soil_2017] identified the anthropogenic SOC debt for the first meter of the soil profile due to land cover change at around 116 GtC (37 GtC for the first 30 cm), compared to previous estimates of 60--130 GtC for the first meter [@lal_world_2001].

Global assessments of the carbon cycle via dynamic global vegetation models (DGVMs) and Earth System Models (ESMs) or bookkeeping models (BKMs) have analyzed SOC losses as part of a comprehensive evaluation of the global carbon budget and land-use change (LUC) emissions [@friedlingstein_global_2019]. While providing estimates of the magnitude of SOC losses due to land cover change, most models lack a detailed consideration of agricultural management. Earlier DGVM and ESM based assessments only considered changes in land cover, but ignored the removal of biomass at harvest [@strassmann_simulating_2008; @betts_climate_2015]. BKMs are designed to estimate LUC related emissions but often ignore changes in SOC due to climate change, CO~2~ fertilization, N deposition. Whereas BKM largely improved in estimating additional emission from wood harvest and shifting cultivation, state of the art models do not consider impacts of varying agricultural management [@friedlingstein_global_2019; @houghton_carbon_2012; @hansis_relevance_2015] (other citations).

@@

Field-scale models [@del_grosso_simulated_2001, @coleman_simulating_1997, @smith_estimating_2010, @taghizadeh-toosi_c-tool_2014] are able to better account for historic agricultural management if detailed information on crop yield levels, fertilizer inputs and various other on-farm measures is available for the studied sites. However, due to the lack of comprehensive global management data as input to these models, scaling up to the global domain remains a complex challenge [@morais_detailed_2019].

However, managed soil of the world have been increasingly studied not only for their carbon emitting behavior, but also because of their capacity to re-store carbon (very long list of citations). These soil carbon sequestration (SCS) techniques have been proposed together with other nature-based solution as a mitigation and negative emission technique (NET) to combat climate change. Technical SCS potentials have been estimated by (1) up-scaling measured SCS potentials from field-scale experiments (cite), or (2) modeling SCS potentials using stylized future scenarios (cite). They range between 0.1 -- 3 GtC per year (cite) with best estimates at 0.5 -- 1.4 GtC per year. Assuming SCS application from 2030 till 2100, this accumulates to 35 -- 98 GtC sequestered again, which is just slightly less than the estimated for total, global SOC debt. Assessing a more realistic potential considering the interdependency with environmental, social and economic sustainability targets has been difficult so far, as also integrated assessment models (cite) have not been able to track management impacts on SOC stocks so far.

The objective of our study is to provide global as well as spatially explicit SOC and SOC debt estimates that considers spatially explicit and time-variant agricultural management. To achieve this objective we create and provide a comprehensive data set of the global gridded management time series data, including crop production levels, residue returning rates, manure application, and the adoption of irrigation and tillage practices. We simulate SOC stocks, dynamics and the SOC debt for 1975--2010 using a yearly, spatial explicit, reduced-complexity SOC model. We decompose the contribution of different management activities through a scenario analysis, identifying the most impacting management decisions for SOC development. Moreover, we compare our model performance against other SOC stock and SOC emission estimates, to evaluate suitability of this approach for further research.

### 2021-08-09 SOCbudgetPaper/methods.Rmd
udate paper draft - intro, abstract, small changes in methods

We calculate $SOC$ stocks based on the area shares of land-use types ($lu$) within our grid cells ($i$). If land is converted from one land-use type $lu=\{crop,natveg\}$ into the other $!lu=\{natveg,crop\}$, a respective share of the $SOC$ is reallocated. We account for land conversion at the beginning of each time step $t$ by calculating a preliminary stock $SOC_{t^*}$ via

@@

To initialize all SOC sub-pools we assume that cropland and natural vegetation are in a land-use type specific steady state for the initialization year 1901. While land-use, climate and litter input information are available from 1901, agricultural management data on residue and manure inputs are held constant at the level of 1965. After a spin-up period of 64 years (1901-1965) with constant agricultural input data, we run the model for additional 10 years with historic input data and start analyzing results from 1975 onwards. Irrigation areas are part of the land-use data set and therefore dynamic from 1901 on, whereas data on no-tillage area is only available after 1974.

>>>>

We calculate $SOC$ stocks based on the area shares of land-use types ($lu$) within our grid cells ($i$). If land is converted from one land-use type $lu=\{crop,natveg\}$ into the other $!lu=\{natveg,crop\}$, a respective share of the $SOC$ is reallocated within our budget. As we do not distinguish between freshly converted and old cropland (see \@ref(fig:lutransitions)), we update the average SOC stock per area within this modeling step. We account for land conversion at the beginning of each time step $t$ by calculating a preliminary stock $SOC_{t^*}$ via

@@

We initialize our SOC sub-pools using a three-step approach, since input data availability is limited for climate and litter (from 1901) as well as for agricultural management data (from 1965):

Firstly, to account for the impacts of legacy fluxes from land-use changes long before the time horizon of interest, we account for land-use change from 1510. We repeat climate and subsequently also litter input data from 1901-1930 to simulate a constant climate for this first initializing period. Agricultural management data is hold constant at the level of 1965. In 1510 we assume all SOC pools to be in natural steady-state, implicating that all land-use change emissions prior to that time, are redistributed to 1510. Due to this long first spin-up period we suppose that the effect of this late accounting and appertaining delayed legacy fluxes can be neglected in 1901. This approaches follows others studies looking on effects of land-use change and management (cite LPJmL (which?)).

Secondly, to account for the impact of a changing climate, we account not only for land-use change, but also for historical climate and consequently natural litter inputs to the soil, from 1901 to 1965 still considering constant agricultural input data.

Thirdly, we run the model for additional 10 years with historic dynamic input data and start analyzing results from 1975 onward. This is in line with IPCC method suggestion to have a 5-20 year spin-up period.

Irrigation areas are part of the water effect associated with the climate data set and therefore dynamic from 1901 on, whereas data on no-tillage areas is only available after 1974.
